---
title: "Machine Learning with Network Data"
subtitle: "Networks I: Concepts and Algorithms | BSE Data Science"
author: "Michael Fehl"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    number-sections: true
    code-fold: false
    theme: flatly
    embed-resources: false
  pdf:
    toc: true
    number-sections: true
execute:
  warning: false
  message: false
---

```{r}
#| label: setup
#| echo: false
library(igraph)
library(Matrix)
library(R.matlab)
library(ggplot2)
library(dplyr)

theme_set(
  theme_classic(base_size = 11) +
    theme(plot.title = element_text(face = "bold"),
          axis.line.y = element_blank(),
          axis.ticks.y = element_blank())
)

# NMI helper (pure R, no sklearn)
nmi_r <- function(a, b) {
  tab   <- table(a, b)
  n_obs <- sum(tab)
  p_xy  <- tab / n_obs
  p_x   <- rowSums(p_xy)
  p_y   <- colSums(p_xy)
  H_x   <- -sum(p_x[p_x > 0] * log(p_x[p_x > 0]))
  H_y   <- -sum(p_y[p_y > 0] * log(p_y[p_y > 0]))
  H_xy  <- -sum(p_xy[p_xy > 0] * log(p_xy[p_xy > 0]))
  MI    <- H_x + H_y - H_xy
  if ((H_x + H_y) == 0) return(0)
  2 * MI / (H_x + H_y)
}

# Load American75 for recurring examples
mat      <- readMat("American75.mat")
A_sp     <- mat$A
A_t      <- as(A_sp, "TsparseMatrix")
edges_fb <- cbind(A_t@i + 1L, A_t@j + 1L)
g_fb     <- graph_from_edgelist(edges_fb, directed = FALSE)
g_fb     <- simplify(g_fb)
```

# What is ML with Network Data?

## Shifting from Generative to Inferential Questions

In Lectures 1 and 2 we built tools for *describing* and *generating* networks.
We now shift to **inference**: given an observed network (possibly with node and
edge attributes), what can we learn *from* it?

Three classes of inference problem:

| Task | Input | Output |
|------|-------|--------|
| **Community detection** | Graph $G$ | Partition of $V$ into groups |
| **Node classification** | Graph $G$, partial labels $y_i$ | Labels for unlabelled nodes |
| **Link prediction** | Graph $G$ (incomplete) | Probability of unobserved edges |

And one representation problem underlying all of them:

| **Node embedding** | Graph $G$ (+ features) | Vectors $z_i \in \mathbb{R}^d$ |

These are not independent problems. A good embedding makes node
classification and link prediction trivial (linear classifier or dot product).
Community detection can be cast as clustering in embedding space.

## A Note on Network Incompleteness

Real network data is almost always **incomplete** in multiple ways:

- **Missing edges**: people do not report all friendships; surveys have recall
  bias toward recent, high-status, or same-gender contacts
- **Missing node attributes**: survey non-response, privacy restrictions
- **Missing nodes entirely**: only a slice of the population is observed

This incompleteness is not random: it is systematically correlated with the
network structure itself. Methods that assume missing-at-random can be badly
misleading.

---

# Community Detection

## The Problem: Finding Latent Groups

A **community** is a set of nodes that are *more densely connected internally*
than to the rest of the graph. This is an intuitive but deliberately vague
definition.

Why do we care?

- **Market segmentation**: find groups of consumers with similar behaviour
- **Political analysis**: identify ideological blocs in voting or communication networks
- **Biological networks**: find functional modules in protein interaction networks
- **Policy**: identify who to target with interventions (information, vaccines,
  microfinance)

## Modularity: A Principled Score

The standard measure of community quality is **modularity** $Q$
(Newman & Girvan 2004). The idea: compare the fraction of edges within
communities to what we would expect in the *configuration model* (same degree
sequence, random wiring).

$$Q = \frac{1}{2m} \sum_{i,j} \left(A_{ij} - \frac{k_i k_j}{2m}\right) \mathbb{1}[c(i) = c(j)]$$

Equivalently, summing over communities $c$:

$$Q = \sum_c \left[\frac{m_c}{m} - \left(\frac{\kappa_c}{2m}\right)^2\right]$$

where $m_c$ = edges within community $c$, and $\kappa_c = \sum_{i \in c} k_i$
= total degree of community $c$.

**Interpretation**: the first term is the actual fraction of edges inside the
community; the second is the fraction expected at random. A good partition has
$Q$ close to 1 (all edges internal), a random partition has $Q \approx 0$.

::: {.callout-note}
## Why the configuration model as null?
The configuration model benchmark (not ER) is the right null because it controls
for degree heterogeneity. In ER, high-degree nodes automatically "look like"
hub communities; using ER as null would find spurious communities wherever
there are hubs. Conditioning on degree removes this confound.
:::

## The Louvain Algorithm

Finding the partition maximising $Q$ is NP-hard. The **Louvain method**
(Blondel et al. 2008) is the standard greedy approximation:

**Phase 1** (local optimisation): Start with each node in its own community.
For each node $i$, compute the $\Delta Q$ gain from moving $i$ into each
neighbouring community:

$$\Delta Q = \frac{k_{i,\text{in}}}{m} - \frac{k_i \Sigma_{\text{tot}}}{2m^2}$$

where $k_{i,\text{in}}$ = edges from $i$ to the target community,
$\Sigma_{\text{tot}}$ = total degree of the target community.
Move $i$ to the community with largest positive $\Delta Q$. Repeat
until no improvement exists.

**Phase 2** (coarsening): Collapse each community into a single super-node,
aggregate edge weights. Repeat Phase 1 on this coarser graph.

**Complexity**: $O(n \log n)$ in practice. Works on networks with
millions of nodes.

::: {.callout-warning}
## Resolution limit
Modularity has a fundamental flaw: it cannot detect communities smaller
than $\sqrt{m}$ edges, because merging two small communities can increase $Q$
even if they have no internal connections. For networks with $m = 10^6$,
communities smaller than $\sim 1000$ edges are invisible to any
modularity-based method.

The **Leiden algorithm** (Traag, Waltman & van Eck 2019) fixes a related
bug in Louvain (communities can become internally disconnected) and uses
a refined local search. For serious community detection work, prefer Leiden.
:::

```{r}
#| label: community-detection
#| code-fold: true
#| fig-cap: "Community detection on a planted partition SBM. Louvain vs Leiden vs ground truth."

set.seed(42)

k_comm <- 4L
n_per  <- 60L
n_tot  <- k_comm * n_per
p_in   <- 0.20
p_out  <- 0.02

pref_mat    <- matrix(p_out, nrow = k_comm, ncol = k_comm)
diag(pref_mat) <- p_in
g_sbm       <- sample_sbm(n_tot, pref = pref_mat,
                           block.sizes = rep(n_per, k_comm))
true_labels <- rep(seq_len(k_comm) - 1L, each = n_per)  # 0-indexed to match Python

# Louvain
louvain_mem <- membership(cluster_louvain(g_sbm)) - 1L
# Leiden (requires leidenAlg or igraph >= 1.3 which wraps it)
# Use cluster_leiden if available, fall back to Louvain
leiden_mem <- tryCatch(
  membership(cluster_leiden(g_sbm)) - 1L,
  error = function(e) {
    message("cluster_leiden not available; using Louvain as proxy.")
    louvain_mem
  }
)

nmi_louvain <- nmi_r(true_labels, louvain_mem)
nmi_leiden  <- nmi_r(true_labels, leiden_mem)

cat(sprintf("Community detection on n=%d, k=%d SBM (p_in=%.2f, p_out=%.2f):\n",
            n_tot, k_comm, p_in, p_out))
cat(sprintf("  Louvain:  NMI = %.4f  (communities found: %d)\n",
            nmi_louvain, length(unique(louvain_mem))))
cat(sprintf("  Leiden:   NMI = %.4f  (communities found: %d)\n",
            nmi_leiden,  length(unique(leiden_mem))))
cat("  (NMI = 1.0 is perfect; NMI = 0 is no better than random)\n")

Q_louvain <- modularity(g_sbm, louvain_mem + 1L)
Q_leiden  <- modularity(g_sbm, leiden_mem  + 1L)
cat(sprintf("\n  Q (Louvain) = %.4f\n", Q_louvain))
cat(sprintf("  Q (Leiden)  = %.4f\n",  Q_leiden))

# Sweep eps (community sharpness)
set.seed(7)
eps_range <- seq(0, 2 * sqrt(k_comm * p_in * n_tot), length.out = 20)
nmis_louvain <- numeric(length(eps_range))
nmis_leiden  <- numeric(length(eps_range))

for (i in seq_along(eps_range)) {
  eps     <- eps_range[i]
  p_in_v  <- min(1, (2 * 4 + eps) / (2 * n_tot))
  p_out_v <- max(0, (2 * 4 - eps) / (2 * n_tot))
  pm_v    <- matrix(p_out_v, k_comm, k_comm); diag(pm_v) <- p_in_v
  g_v     <- sample_sbm(n_tot, pref = pm_v, block.sizes = rep(n_per, k_comm))
  mem_l   <- membership(cluster_louvain(g_v)) - 1L
  mem_d   <- tryCatch(membership(cluster_leiden(g_v)) - 1L,
                      error = function(e) mem_l)
  nmis_louvain[i] <- nmi_r(true_labels, mem_l)
  nmis_leiden[i]  <- nmi_r(true_labels, mem_d)
}

par(mar = c(4, 4, 4, 1))
plot(eps_range, nmis_louvain,
     type = "b", pch = 19, cex = 0.7, col = "#4C72B0", lwd = 2,
     xlab = expression(epsilon~~"(community sharpness)"),
     ylab = "NMI with ground truth",
     main = "Community Detection Quality vs. Signal Strength",
     ylim = c(-0.02, 1.05))
lines(eps_range, nmis_leiden, type = "b", pch = 15, cex = 0.7,
      col = "#C44E52", lwd = 2)
abline(v = 2 * sqrt(k_comm * p_in * n_tot) * 0.5,
       col = "gray", lty = 2, lwd = 1.5)
legend("topleft",
       legend = c("Louvain", "Leiden", "Approx. detectability threshold"),
       col    = c("#4C72B0", "#C44E52", "gray"),
       lwd    = 2, pch = c(19, 15, NA), lty = c(1, 1, 2),
       bty    = "n", cex = 0.9)
```

## SBM Inference as an Alternative

Modularity optimisation lacks a generative model; we are maximising a score,
not fitting a probability model. The **SBM inference** approach (from Lecture 2)
is more principled:

1. Specify a generative model ($k$ communities, mixing matrix $M$)
2. Write down the likelihood $\mathcal{L}(G \mid z, M)$
3. Maximise over both $z$ and $M$; MLE gives $\hat{M}_{uv} = E_{uv}/N_{uv}$
4. Use BIC to choose $k$: $\text{BIC} = -2 \ln \hat{\mathcal{L}} + p \ln n$
   where $p = k + \binom{k}{2}$ parameters

**Advantage**: principled model selection, well-calibrated uncertainty,
connects to the detectability threshold theory.

**Disadvantage**: the likelihood landscape is non-convex; many local optima
exist. Run multiple random initialisations and select the best.

---

# Node Classification and Label Propagation

## The Semi-Supervised Setup

Given a graph $G = (V, E)$ where a *small* subset $V_L \subset V$ has known
labels $y_i \in \{1, \ldots, K\}$, predict labels for $V_U = V \setminus V_L$.

The key assumption here is **homophily**: connected nodes tend to share labels.
This holds in social networks (birds of a feather), citation networks (papers
cite related papers), and many biological networks.

::: {.callout-warning}
## When homophily fails
Fraud detection networks often exhibit **heterophily**: fraudsters connect to
legitimate users in order to appear trustworthy. In such networks, label
propagation gives the wrong answer; a node surrounded by legitimate users
is *more* suspicious, not less. Always validate the homophily assumption.
:::

## Label Propagation (Zhu & Ghahramani 2002)

The simplest and most robust algorithm:

**Initialisation**: $F_i = y_i$ for labelled nodes (one-hot), uniform for unlabelled.

**Update rule** (iterative):
$$F_i \leftarrow \frac{1}{k_i} \sum_{j \in \mathcal{N}(i)} F_j, \quad \forall i \in V_U$$

**Clamping**: after each step, reset labelled nodes to their true values.

**Convergence**: iterate until $\|F^{(t+1)} - F^{(t)}\|_\infty < \epsilon$.

In matrix form, this is a fixed-point iteration on the random-walk matrix
$D^{-1} A$. The closed-form solution is:

$$F_U = (I - D_{UU}^{-1} A_{UU})^{-1} D_{UU}^{-1} A_{UL} Y_L$$

**Why it works**: the random-walk matrix $D^{-1}A$ is the transition matrix
of a lazy random walk. Each update propagates label distributions one step
along random walks. Under homophily, walks starting from a labelled node tend
to stay in the same community and spread the correct label.

**Complexity**: $O(m \cdot K \cdot T)$ where $T$ is the number of iterations
(typically 20–100). Very fast for sparse graphs.

```{r}
#| label: label-propagation
#| code-fold: true
#| fig-cap: "Label propagation accuracy vs. fraction labelled, on planted partition SBM."

label_propagation <- function(g, Y, labeled_mask, max_iter = 200, tol = 1e-8) {
  # Y           : (n x K) numeric matrix, one-hot for labelled nodes
  # labeled_mask: logical vector length n
  n    <- vcount(g)
  K    <- ncol(Y)
  F    <- Y
  nbrs <- lapply(seq_len(n), function(v) neighbors(g, v))
  degs <- pmax(sapply(nbrs, length), 1L)

  for (iter in seq_len(max_iter)) {
    F_old <- F
    for (v in seq_len(n)) {
      if (labeled_mask[v] || length(nbrs[[v]]) == 0) next
      F[v, ] <- colSums(F[nbrs[[v]], , drop = FALSE]) / degs[v]
    }
    F[labeled_mask, ] <- Y[labeled_mask, ]
    if (max(abs(F - F_old)) < tol) break
  }
  F
}

set.seed(0)
n_per_c  <- 40L; k_c <- 3L; n_c <- n_per_c * k_c
pref_c   <- matrix(0.02, k_c, k_c); diag(pref_c) <- 0.3
g_c      <- sample_sbm(n_c, pref = pref_c, block.sizes = rep(n_per_c, k_c))
true_labels_c <- rep(seq_len(k_c) - 1L, each = n_per_c)   # 0-indexed

label_fracs  <- c(0.01, 0.03, 0.05, 0.10, 0.20, 0.30, 0.50)
acc_lp_list  <- numeric(length(label_fracs))
acc_lr_list  <- numeric(length(label_fracs))

# Simple multinomial logistic (degree-only, no graph)
multinomial_lr <- function(X_train, y_train, X_test, K) {
  # One-vs-rest logistic regression via glm
  preds <- matrix(0, nrow = NROW(X_test), ncol = K)
  for (k in seq_len(K)) {
    df_train <- data.frame(y = as.integer(y_train == (k - 1L)), x = X_train)
    fit <- glm(y ~ x, data = df_train, family = binomial())
    preds[, k] <- predict(fit, newdata = data.frame(x = X_test), type = "response")
  }
  apply(preds, 1, which.max) - 1L
}

deg_feat <- degree(g_c)

for (fi in seq_along(label_fracs)) {
  frac         <- label_fracs[fi]
  labeled_mask <- logical(n_c)
  for (cc in seq_len(k_c)) {
    idx     <- which(true_labels_c == (cc - 1L))
    n_pick  <- max(1L, as.integer(length(idx) * frac))
    labeled_mask[sample(idx, n_pick)] <- TRUE
  }

  # Label propagation
  Y <- matrix(0, n_c, k_c)
  for (i in seq_len(n_c)) Y[i, true_labels_c[i] + 1L] <- 1.0
  F0 <- Y; F0[!labeled_mask, ] <- 1.0 / k_c
  F_pred <- label_propagation(g_c, F0, labeled_mask)
  pred_lp <- apply(F_pred, 1, which.max) - 1L
  acc_lp_list[fi] <- mean(pred_lp[!labeled_mask] == true_labels_c[!labeled_mask])

  # LR on degree only
  pred_lr <- multinomial_lr(deg_feat[labeled_mask], true_labels_c[labeled_mask],
                             deg_feat[!labeled_mask], k_c)
  acc_lr_list[fi] <- mean(pred_lr == true_labels_c[!labeled_mask])
}

par(mar = c(4, 4, 4, 1))
plot(label_fracs * 100, acc_lp_list,
     type = "b", pch = 19, cex = 0.8, col = "#C44E52", lwd = 2,
     xlab = "% nodes labelled", ylab = "Accuracy on unlabelled nodes",
     main = sprintf("Label Propagation vs. No-Graph Baseline\n(n=%d, k=%d communities)",
                    n_c, k_c),
     ylim = c(0, 1.05))
lines(label_fracs * 100, acc_lr_list, type = "b", pch = 15, cex = 0.8,
      col = "#4C72B0", lwd = 2, lty = 2)
abline(h = 1 / k_c, col = "gray", lty = 3, lwd = 1.5)
legend("bottomright",
       legend = c("Label Propagation (uses graph)",
                  "LR on degree only (no graph)",
                  sprintf("Random baseline (1/%d)", k_c)),
       col    = c("#C44E52", "#4C72B0", "gray"),
       lwd    = 2, lty = c(1, 2, 3), pch = c(19, 15, NA),
       bty    = "n", cex = 0.9)

frac_idx <- which(label_fracs == 0.10)
cat(sprintf("Accuracy at 10%% labelled:\n"))
cat(sprintf("  Label Propagation: %.4f\n", acc_lp_list[frac_idx]))
cat(sprintf("  LR (no graph):     %.4f\n", acc_lr_list[frac_idx]))
```

## When to Use What: A Decision Ladder

Before reaching for a GNN, always try the simpler methods first:

1. **Majority class baseline**: does your "prediction" just rediscover the
   class imbalance? Always compute this.
2. **LR on node features only**: how much does the graph add, beyond features?
3. **Label propagation (graph only)**: how much do features add, beyond graph?
4. **ICA (features + iterated neighbor statistics)**: combine both without
   a neural network.
5. **GCN / GraphSAGE**: use when large labelled datasets exist and the
   gain from steps 1–4 is insufficient.

Steps 1–4 are interpretable, fast, and usually get 80–90% of GNN performance.
The marginal gain from GNNs is often small on homophilic networks with
clean node features.

---

# Link Prediction

## Problem Setup

Given the observed graph $G = (V, E)$, predict which non-edges $(i, j) \notin E$
are "true" edges; either missing from our measurement, or likely to form in
the future.

**Evaluation protocol**:
1. Hold out 10–20% of edges as *positive* test examples
2. Sample an equal number of *negative* examples (non-edges)
3. Score all candidate pairs with a similarity function $s(i, j)$
4. Compute AUC-ROC: $\Pr[s(\text{true edge}) > s(\text{non-edge})]$

**Negative sampling matters**: random non-edges are too easy (most pairs have
$s \approx 0$). Hard negatives at distance 2 or 3 give more realistic evaluation.

## Structural Similarity Scores

All heuristics score $(i, j)$ based on their common neighbourhood $\mathcal{N}(i) \cap \mathcal{N}(j)$.

**Common Neighbours (CN)**:
$$s_{\text{CN}}(i,j) = |\mathcal{N}(i) \cap \mathcal{N}(j)|$$

**Jaccard coefficient** (normalises for hub effect):
$$s_J(i,j) = \frac{|\mathcal{N}(i) \cap \mathcal{N}(j)|}{|\mathcal{N}(i) \cup \mathcal{N}(j)|}$$

**Adamic–Adar** (Adamic & Adar 2003): down-weight common neighbours
with high degree (high-degree nodes are "common friends" who are friends
with everyone and therefore carry less information):

$$s_{\text{AA}}(i,j) = \sum_{w \in \mathcal{N}(i) \cap \mathcal{N}(j)} \frac{1}{\log k_w}$$

**Resource Allocation**: stricter version, using $1/k_w$ instead of $1/\log k_w$.

**Preferential Attachment**: no common-neighbour computation needed, just:
$$s_{\text{PA}}(i,j) = k_i \cdot k_j$$

PA is useful in scale-free networks: hubs grow by attaching to other hubs.
It has $O(1)$ computation per pair and scales to very large graphs.

## Global Similarity: Katz Index

Common-neighbour methods only look one hop away. The **Katz index**
aggregates over *all* path lengths, weighted by a decay factor $\beta < 1$:

$$K(i,j) = \sum_{l=1}^{\infty} \beta^l [A^l]_{ij} = [(I - \beta A)^{-1} - I]_{ij}$$

$[A^l]_{ij}$ counts the number of walks of length $l$ from $i$ to $j$.
The $\beta$ discount ensures convergence (requires $\beta < 1/\lambda_{\max}(A)$).

Matrix inversion is $O(n^3)$, so for large graphs we approximate: run
$L$ steps of the power iteration $(I - \beta A)^{-1} \approx \sum_{l=0}^L \beta^l A^l$
using sparse matrix-vector multiplications ($O(Lm)$).

## Measuring Performance: The AUC

The **ROC curve** plots the true positive rate (TPR) against the false positive
rate (FPR) as the score threshold varies. **AUC** is the area under this curve:

$$\text{AUC} = \Pr[s(\text{true edge}) > s(\text{random non-edge})]$$

AUC = 0.5: no better than random. AUC = 1.0: perfect separation.

```{r}
#| label: link-prediction
#| code-fold: true
#| fig-cap: "Link prediction AUC for various heuristics on American75 Facebook network."

# ---- Similarity functions ----
cn_score <- function(g, u, v) {
  length(intersect(neighbors(g, u), neighbors(g, v)))
}

jaccard_score <- function(g, u, v) {
  nu <- neighbors(g, u); nv <- neighbors(g, v)
  denom <- length(union(nu, nv))
  if (denom == 0) return(0.0)
  length(intersect(nu, nv)) / denom
}

adamic_adar <- function(g, u, v) {
  common <- intersect(neighbors(g, u), neighbors(g, v))
  if (length(common) == 0) return(0.0)
  d_w <- degree(g, common)
  sum(ifelse(d_w > 1, 1 / log(d_w), 0))
}

pref_attach <- function(g, u, v) {
  degree(g, u) * degree(g, v)
}

# ---- AUC helper ----
auc_roc <- function(y_true, y_score) {
  # Wilcoxon-Mann-Whitney estimator: P(score_pos > score_neg)
  pos <- y_score[y_true == 1]
  neg <- y_score[y_true == 0]
  if (length(pos) == 0 || length(neg) == 0) return(NaN)
  mean(outer(pos, neg, ">")) + 0.5 * mean(outer(pos, neg, "=="))
}

evaluate_lp <- function(g, score_fn, n_test = 500L, seed = 42L) {
  set.seed(seed)
  edges    <- as_edgelist(g, names = FALSE)  # (m x 2) matrix, 1-indexed
  n        <- vcount(g)
  n_test   <- min(n_test, nrow(edges) %/% 5L)
  test_idx <- sample(nrow(edges), n_test)
  test_pos <- edges[test_idx, , drop = FALSE]

  # Train graph: remove test edges
  train_g <- delete_edges(g, get.edge.ids(g, t(test_pos)))

  # Build edge set for fast lookup
  edge_set <- apply(edges, 1, function(r) paste(sort(r), collapse = "_"))
  edge_set  <- as.environment(as.list(setNames(rep(TRUE, length(edge_set)), edge_set)))

  # Sample negatives
  negs  <- matrix(0L, 0L, 2L)
  tries <- 0L
  while (nrow(negs) < n_test && tries < n_test * 200L) {
    u <- sample.int(n, 1L); v <- sample.int(n, 1L)
    key <- paste(sort(c(u, v)), collapse = "_")
    if (u != v && !exists(key, envir = edge_set, inherits = FALSE))
      negs <- rbind(negs, c(u, v))
    tries <- tries + 1L
  }

  y_true  <- c(rep(1L, n_test), rep(0L, nrow(negs)))
  y_score <- c(
    apply(test_pos, 1, function(r) score_fn(train_g, r[1], r[2])),
    apply(negs,     1, function(r) score_fn(train_g, r[1], r[2]))
  )
  auc_roc(y_true, y_score)
}

methods <- list(
  list(name = "Common Neighbours", fn = cn_score),
  list(name = "Jaccard",           fn = jaccard_score),
  list(name = "Adamic-Adar",       fn = adamic_adar),
  list(name = "Pref. Attachment",  fn = pref_attach)
)

cat("Link prediction AUC (American75, n_test=500):\n")
aucs <- setNames(numeric(length(methods)), sapply(methods, `[[`, "name"))
for (m in methods) {
  auc <- evaluate_lp(g_fb, m$fn, n_test = 500L, seed = 0L)
  aucs[m$name] <- auc
  bar <- paste(rep("\u2588", as.integer(auc * 40)), collapse = "")
  cat(sprintf("  %-24s AUC = %.4f  %s\n", m$name, auc, bar))
}
cat("  Random baseline:       AUC = 0.5000\n")
cat("\nAdamic-Adar typically outperforms CN and Jaccard because it\n")
cat("down-weights high-degree 'universal friends' who provide less information.\n")

# Plot
par(mar = c(4, 10, 4, 2))
all_names <- c(names(aucs), "Random")
all_aucs  <- c(aucs, Random = 0.5)
bar_cols  <- c(rep("#4C72B0", length(aucs)), "#aaaaaa")
bp <- barplot(all_aucs, horiz = TRUE, names.arg = all_names,
              col = bar_cols, border = NA, las = 1,
              xlab = "AUC-ROC",
              main = "Link Prediction on American75 Facebook Network",
              xlim = c(0, 1.1))
abline(v = 0.5, col = "red", lty = 2, lwd = 1.5)
text(all_aucs + 0.01, bp, sprintf("%.3f", all_aucs), cex = 0.85, adj = 0)
```

---

# Node Embeddings: Mapping Structure to Vectors

## Why Embeddings?

Structural similarity scores like Adamic–Adar are hand-crafted and task-specific.
A more flexible approach: learn a mapping $\text{ENC}: V \to \mathbb{R}^d$
such that structurally similar nodes have similar embeddings.

Then:
- **Link prediction**: $s(i, j) = \sigma(z_i^\top z_j)$
- **Node classification**: logistic regression on $z_i$
- **Community detection**: $k$-means on $\{z_i\}$
- **Visualisation**: 2D PCA or UMAP of $\{z_i\}$

The key design choice is what "similar" means; different definitions lead
to different algorithms.

## Shallow Embeddings: The Matrix Factorization View

**Spectral embedding** (Laplacian Eigenmaps): find vectors that minimise

$$\min_{Z} \sum_{(i,j) \in E} \|z_i - z_j\|^2 \quad \text{s.t. } Z^\top D Z = I$$

The solution is the $d$ smallest non-trivial eigenvectors of the normalised
Laplacian $L_{\text{sym}} = D^{-1/2}(D-A)D^{-1/2}$.

This is equivalent to truncated SVD on $A_{\text{sym}} = D^{-1/2} A D^{-1/2}$.
The top-$d$ singular vectors of the normalised adjacency give embeddings
where dot products approximate random-walk co-occurrence probabilities.

**Limitation**: these are *transductive*; the embedding is fixed for the
training graph. A new node that joins the network needs to be re-embedded.

## DeepWalk: Random Walks as Sentences

**Perozzi et al. (2014)** drew an analogy between graphs and text:
nodes are words, random walks are sentences, co-occurrence in a walk
window is analogous to co-occurrence in a text context window.

**Algorithm**:

1. For each node $v$, generate $r$ random walks of length $\ell$ starting at $v$
2. Treat each walk as a "sentence" of node IDs
3. Run **skip-gram** with negative sampling: maximise
   $$\sum_{v \in V} \sum_{u \in \mathcal{N}_R(v)} \log \sigma(z_v^\top z_u)
   + \sum_{k=1}^K \mathbb{E}_{u_k \sim P_n} [\log \sigma(-z_v^\top z_{u_k})]$$

where $\mathcal{N}_R(v)$ is the multi-set of nodes in the walk neighbourhood
of $v$ (within window $w$), and $P_n$ is the noise distribution for negative
sampling.

**Why it works**: nodes that frequently co-occur on short random walks;
meaning they are in the same community or tightly connected subgraph;
get assigned similar embedding vectors. The skip-gram objective
*factorises* the PMI (pointwise mutual information) matrix of co-occurrences.

## Node2Vec: Interpolating BFS and DFS

DeepWalk uses uniform random walks. **Node2Vec** (Grover & Leskovec 2016)
introduces a 2nd-order Markov bias controlled by two parameters:

- **Return parameter $p$**: probability of returning to the previous node
  (inversely proportional to $1/p$). High $p$ → less backtracking.
- **In-out parameter $q$**: controls BFS vs DFS. $q < 1$ → DFS (explore
  outward), $q > 1$ → BFS (stay local).

At node $v$, having arrived from node $u$, transition to $x$ with unnormalised probability:

$$\alpha_{pq}(u, x) = \begin{cases} 1/p & d(u, x) = 0 \;\text{(return)} \\ 1 & d(u, x) = 1 \;\text{(lateral)} \\ 1/q & d(u, x) = 2 \;\text{(outward)} \end{cases}$$

**Why this matters**:

- **BFS walks** ($q \gg 1$): explore the local neighbourhood. Nodes with
  similar local structure (same community) get similar embeddings.
  Captures **homophily**.
- **DFS walks** ($q \ll 1$): explore outward, sampling diverse parts of
  the graph. Nodes at the same *role* in different communities (e.g., all
  hub nodes) get similar embeddings. Captures **structural equivalence**.

```{r}
#| label: node-embeddings
#| code-fold: true
#| fig-cap: "Spectral embedding of an SBM. Communities cluster clearly in 2D."

set.seed(0)
k_emb  <- 4L; n_e <- 30L; n_emb <- k_emb * n_e
pref_e <- matrix(0.02, k_emb, k_emb); diag(pref_e) <- 0.3
g_emb  <- sample_sbm(n_emb, pref = pref_e, block.sizes = rep(n_e, k_emb))
true_labels_e <- rep(seq_len(k_emb) - 1L, each = n_e)

# Build normalised adjacency A_norm = D^{-1/2} A D^{-1/2}
A_e     <- as_adjacency_matrix(g_emb, sparse = TRUE)
deg_e   <- Matrix::rowSums(A_e)
d_isqrt <- ifelse(deg_e > 0, 1 / sqrt(deg_e), 0)
D_isqrt <- Diagonal(x = d_isqrt)
A_norm  <- D_isqrt %*% A_e %*% D_isqrt

# Top-5 singular vectors via RSpectra (or base svd on small graph)
# For small n use base svd; for large use RSpectra::svds
svd_res <- if (requireNamespace("RSpectra", quietly = TRUE)) {
  RSpectra::svds(A_norm, k = 5L)
} else {
  sv <- svd(as.matrix(A_norm), nu = 5L, nv = 5L)
  list(u = sv$u, d = sv$d, v = sv$v)
}
# Skip leading (trivial) eigenvector; take cols 2 and 3
emb <- svd_res$u[, 2:3, drop = FALSE]

# Random walk helper
random_walk <- function(g, start, len) {
  walk <- integer(len)
  walk[1] <- start
  for (i in seq_len(len - 1)) {
    nb <- neighbors(g, walk[i])
    if (length(nb) == 0) { walk <- walk[seq_len(i)]; break }
    walk[i + 1] <- sample(nb, 1L)
  }
  walk
}

par(mfrow = c(1, 2), mar = c(4, 4, 4, 1))

## (A) 2D spectral embedding
emb_cols <- c("#4C72B0", "#C44E52", "#55A868", "#DD8452")
plot(NULL, xlim = range(emb[, 1]) * 1.1, ylim = range(emb[, 2]) * 1.1,
     xlab = "Eigenvector 2", ylab = "Eigenvector 3",
     main = "2D Spectral Embedding of SBM\n(communities cluster by construction)")
for (cc in seq_len(k_emb)) {
  mask_c <- true_labels_e == (cc - 1L)
  points(emb[mask_c, 1], emb[mask_c, 2],
         pch = 19, cex = 0.9, col = adjustcolor(emb_cols[cc], 0.8))
}
legend("topright", legend = paste0("Community ", 0:(k_emb - 1)),
       col = emb_cols, pch = 19, bty = "n", cex = 0.85)

## (B) Random walk co-occurrence heatmap (4x4, aggregated by community)
set.seed(42)
n_walks  <- 30L; walk_len <- 15L; window <- 3L
walks    <- lapply(seq_len(n_emb), function(v)
  replicate(ceiling(n_walks / n_emb) + 1L,
            random_walk(g_emb, v, walk_len), simplify = FALSE))
walks    <- unlist(walks, recursive = FALSE)

cooc <- matrix(0, n_emb, n_emb)
for (w in walks) {
  lw <- length(w)
  for (i in seq_len(lw)) {
    js <- seq(max(1L, i - window), min(lw, i + window))
    js <- js[js != i]
    cooc[w[i], w[js]] <- cooc[w[i], w[js]] + 1L
  }
}

comm_cooc <- matrix(0, k_emb, k_emb)
for (u in seq_len(n_emb))
  for (v in seq_len(n_emb))
    comm_cooc[true_labels_e[u] + 1L, true_labels_e[v] + 1L] <-
      comm_cooc[true_labels_e[u] + 1L, true_labels_e[v] + 1L] + cooc[u, v]
comm_cooc <- comm_cooc / rowSums(comm_cooc)

image(t(comm_cooc[k_emb:1, ]), col = hcl.colors(20, "Blues", rev = TRUE),
      xaxt = "n", yaxt = "n",
      xlab = "Landing community", ylab = "Source community",
      main = "Random Walk Co-occurrence\n(diagonal = within-community)")
axis(1, at = seq(0, 1, length.out = k_emb), labels = paste0("C", 0:(k_emb-1)))
axis(2, at = seq(0, 1, length.out = k_emb), labels = paste0("C", (k_emb-1):0), las = 1)

# Spectral embedding similarity: within vs cross community
within_sim <- c(); cross_sim <- c()
for (i in seq_len(n_emb)) {
  for (j in seq_len(n_emb)) {
    if (j <= i) next
    s <- sum(emb[i, ] * emb[j, ])
    if (true_labels_e[i] == true_labels_e[j]) within_sim <- c(within_sim, s)
    else                                       cross_sim  <- c(cross_sim,  s)
  }
}
cat(sprintf("Spectral embedding similarity:\n"))
cat(sprintf("  Within-community: mean = %.4f\n",  mean(within_sim)))
cat(sprintf("  Cross-community:  mean = %.4f\n",  mean(cross_sim)))
cat(sprintf("  Separation ratio: %.2fx\n",
            mean(within_sim) / abs(mean(cross_sim))))
```

## Limitations of Shallow Embeddings

All methods so far produce *transductive* embeddings: one vector per node,
learned on a specific graph. They cannot:

- Embed new nodes without retraining
- Incorporate rich node features (text, demographics, sensor readings)
- Adapt to different downstream tasks (same embedding used for both
  link prediction and community detection)

This motivates **inductive** methods; specifically Graph Neural Networks.

---

# Graph Neural Networks (GNNs)

## The Core Idea: Learned Neighbourhood Aggregation

A GNN replaces the fixed aggregation of shallow methods with a
*learnable* neighbourhood aggregation function:

$$h_v^{(l+1)} = \sigma\!\left(W^{(l)} \cdot \text{AGG}\!\left(\{h_u^{(l)} : u \in \mathcal{N}(v)\}\right) + B^{(l)} h_v^{(l)}\right)$$

where:
- $h_v^{(0)} = x_v$ (node features, or a one-hot vector if none exist)
- $W^{(l)}, B^{(l)}$ are learnable weight matrices, **shared across all nodes**
- AGG is an aggregation function (mean, sum, max, attention-weighted)
- $\sigma$ is a non-linearity (ReLU, etc.)

The sharing of weights across nodes is what makes GNNs **inductive**: the
same function is applied to every node's neighbourhood, so it generalises
to unseen nodes with the same feature distribution.

## Graph Convolutional Network (GCN, Kipf & Welling 2017)

The most widely used variant simplifies to:

$$H^{(l+1)} = \sigma\!\left(\hat{A} H^{(l)} W^{(l)}\right)$$

where $\hat{A} = \tilde{D}^{-1/2} \tilde{A} \tilde{D}^{-1/2}$ is the
symmetrically normalised adjacency with added self-loops
($\tilde{A} = A + I$).

**Why self-loops?** Without them, $H^{(l+1)}_v$ depends only on
neighbours, not on $v$ itself. The self-loop ensures each node
*aggregates its own features* together with its neighbours'.

**Why symmetric normalisation?** It prevents the embeddings of high-degree
nodes from having artificially large magnitudes and makes gradient flow
more stable. It also corresponds to the Laplacian smoothing filter.

**Two-layer GCN for node classification** (fully written out):

$$Z = \text{softmax}\!\left(\hat{A} \cdot \text{ReLU}\!\left(\hat{A} X W^{(0)}\right) W^{(1)}\right)$$

Train by minimising cross-entropy on labelled nodes:
$\mathcal{L} = -\sum_{i \in V_L} \sum_k y_{ik} \ln Z_{ik}$

## GraphSAGE: Scalable Inductive Learning

GCN requires the full adjacency matrix $\hat{A}$ during training. For
a graph with $10^6$ nodes, $\hat{A}$ is a $10^6 \times 10^6$ matrix;
impossible to store or multiply with.

**GraphSAGE** (Hamilton, Ying & Leskovec 2017) introduces *neighbourhood
sampling*: for each node, sample a fixed-size subset of $S$ neighbours
instead of using all of them. This allows mini-batch training:

$$h_v^{(l+1)} = \sigma\!\left(W^{(l)} \left[ h_v^{(l)} \,\|\, \text{AGG}\!\left(\{h_u^{(l)} : u \in \tilde{\mathcal{N}}_S(v)\}\right) \right]\right)$$

(where $\|$ denotes concatenation and $\tilde{\mathcal{N}}_S(v)$ is a
sampled neighbourhood of size $S$).

Because the weight matrices are *shared*, a trained GraphSAGE model
can embed nodes that were not present during training; this is the
inductive property.

## Graph Attention Networks (GAT, Veličković et al. 2018)

Instead of uniform (mean) or fixed-normalisation (GCN) aggregation,
GAT learns **attention weights** $\alpha_{ij}$ for each edge:

$$\alpha_{ij} = \frac{\exp\!\left(\text{LeakyReLU}(a^\top [W h_i \| W h_j])\right)}{\sum_{k \in \mathcal{N}(i)} \exp\!\left(\text{LeakyReLU}(a^\top [W h_i \| W h_k])\right)}$$

$$h_i^{\text{new}} = \sigma\!\left(\sum_{j \in \mathcal{N}(i)} \alpha_{ij} W h_j\right)$$

Attention weights are interpretable: a large $\alpha_{ij}$ means node $i$
relies heavily on node $j$ for its prediction.

**Multi-head attention**: run $K$ independent attention heads and
concatenate (or average) the results. Reduces variance and captures
diverse neighbourhood patterns.

## The Message-Passing Framework

All GNN variants fit a single **message passing neural network (MPNN)** framework
(Gilmer et al. 2017):

1. **Message**: $m_{ij}^{(l)} = \phi\!\left(h_i^{(l)}, h_j^{(l)}, e_{ij}\right)$
2. **Aggregate**: $a_i^{(l)} = \bigoplus_{j \in \mathcal{N}(i)} m_{ij}^{(l)}$
3. **Update**: $h_i^{(l+1)} = \gamma\!\left(h_i^{(l)}, a_i^{(l)}\right)$

The choice of $\phi$, $\bigoplus$, and $\gamma$ defines the GNN variant:

| Model | $\bigoplus$ | Notes |
|-------|------------|-------|
| GCN | Normalised sum | Fixed weights, simple |
| GraphSAGE | Mean / max | Sampled, inductive |
| GAT | Attention-weighted sum | Learnable weights |
| GIN | Sum | Most expressive (WL-equivalent) |

**GIN** (Xu et al. 2019) proves that sum aggregation is necessary for maximum
expressiveness: mean and max aggregations can fail to distinguish different
graph structures that sum can tell apart.

```{r}
#| label: gcn-demo
#| code-fold: true
#| fig-cap: "GCN forward pass (pure R / base matrix ops) demonstrating multi-hop neighbourhood aggregation."

build_a_hat <- function(A_dense) {
  n       <- nrow(A_dense)
  A_tilde <- A_dense + diag(n)
  d       <- rowSums(A_tilde)
  d_isqrt <- ifelse(d > 0, 1 / sqrt(d), 0)
  D_isqrt <- diag(d_isqrt)
  D_isqrt %*% A_tilde %*% D_isqrt
}

gcn_forward <- function(A_hat, X, W1, W2) {
  # Layer 1: ReLU(A_hat %*% X %*% W1)
  H1 <- A_hat %*% X %*% W1
  H1[H1 < 0] <- 0
  # Layer 2 + softmax
  Z  <- A_hat %*% H1 %*% W2
  Z  <- Z - apply(Z, 1, max)          # numerical stability
  Z  <- exp(Z)
  Z  / rowSums(Z)
}

set.seed(42)
k_gcn <- 3L; n_g <- 30L; n_gcn <- k_gcn * n_g
pref_g <- matrix(0.02, k_gcn, k_gcn); diag(pref_g) <- 0.3
g_gcn  <- sample_sbm(n_gcn, pref = pref_g, block.sizes = rep(n_g, k_gcn))
true_labels_g <- rep(seq_len(k_gcn) - 1L, each = n_g)

A_gcn <- as.matrix(as_adjacency_matrix(g_gcn, sparse = FALSE))
A_hat <- build_a_hat(A_gcn)

# Class-informative node features
means_g <- matrix(rnorm(k_gcn * 8) * 2, nrow = k_gcn)
X_gcn   <- means_g[true_labels_g + 1L, ] + matrix(rnorm(n_gcn * 8), n_gcn, 8)

# Random (untrained) weights
W1_gcn <- matrix(rnorm(8 * 16) * 0.1, 8, 16)
W2_gcn <- matrix(rnorm(16 * k_gcn) * 0.1, 16, k_gcn)

Z_gcn <- gcn_forward(A_hat, X_gcn, W1_gcn, W2_gcn)
cat(sprintf("GCN forward pass (untrained weights):\n"))
cat(sprintf("  Input features:    X shape = (%d, %d)\n",  nrow(X_gcn),  ncol(X_gcn)))
cat(sprintf("  After layer 1:     H1 shape = (%d, 16)\n", n_gcn))
cat(sprintf("  Output (logits):   Z shape = (%d, %d)\n",  nrow(Z_gcn),  ncol(Z_gcn)))
cat("\nPredicted class probabilities for first 5 nodes:\n")
for (i in seq_len(5)) {
  cat(sprintf("  Node %d (true class %d): %s\n",
              i - 1L, true_labels_g[i],
              paste(round(Z_gcn[i, ], 3), collapse = "  ")))
}

# Effect of aggregation depth
cat("\n--- Effect of aggregation depth ---\n")
for (depth in 0:2) {
  X_smooth <- X_gcn
  for (d in seq_len(depth)) X_smooth <- A_hat %*% X_smooth
  within_var <- var(X_smooth[true_labels_g == 0, 1])
  cross_var  <- var(sapply(0:(k_gcn - 1), function(cc)
    mean(X_smooth[true_labels_g == cc, 1])))
  cat(sprintf("  Depth %d: within-class var=%.3f, between-class spread=%.3f, ratio=%.2f\n",
              depth, within_var, cross_var, cross_var / max(within_var, 1e-9)))
}
```

## Practical Checklist: When to Use What

```{r}
#| label: gnn-checklist
#| echo: false

df_check <- data.frame(
  Question = c(
    "Homophilic or heterophilic?",
    "Transductive or inductive?",
    "Node features available?",
    "How many GNN layers?",
    "Interpretability needed?",
    "Scale > 1M nodes?",
    "Overlapping communities?"
  ),
  Recommendation = c(
    "Homophilic → GCN/GAT; Heterophilic → GraphSAGE or H2GCN",
    "Transductive → GCN (full graph); Inductive → GraphSAGE",
    "Yes: use them as x_v; No: use degree or identity matrix",
    "2-3 layers (deeper → over-smoothing, all nodes same vector)",
    "GAT (attention weights); GNN-Explainer for post-hoc",
    "Neighbour sampling (GraphSAGE, ClusterGCN, SIGN)",
    "NMF or BigCLAM; hard-partition methods miss overlaps"
  ),
  stringsAsFactors = FALSE
)
print(df_check, right = FALSE, row.names = FALSE)
```

::: {.callout-important}
## Over-smoothing: the depth problem
With $L$ GNN layers, each node aggregates information from its $L$-hop
neighbourhood. For $L \gg 1$, every node's representation converges to
the same vector (proportional to the dominant eigenvector of $\hat{A}$).
This *over-smoothing* destroys the discriminative signal.

In practice: **2 layers suffices for most tasks**. If deeper is needed,
use skip connections (jumping knowledge networks) or residual connections.
:::

---

# Summary: ML on Networks

## The Three-Problem Framework

```{r}
#| label: summary-table
#| echo: false

df_sum <- data.frame(
  Task = c("Community detection", "Node classification",
           "Link prediction",     "Node embedding"),
  Simplest_method = c("Louvain (modularity)", "Label propagation",
                      "Common neighbours",    "Spectral (Laplacian eigenmaps)"),
  Best_ML_method  = c("Leiden / DC-SBM inference", "GCN / GAT",
                      "Katz + GNN (dot product)",  "Node2Vec / GraphSAGE"),
  Key_assumption  = c("Modularity > 0", "Homophily",
                      "Triangle closure", "Proximity ~ similarity"),
  Main_failure    = c("Resolution limit", "Heterophily",
                      "Degree bias",      "Transductive only"),
  stringsAsFactors = FALSE
)
colnames(df_sum) <- c("Task", "Simplest method", "Best ML method",
                      "Key assumption", "Main failure mode")
print(df_sum, right = FALSE, row.names = FALSE)
```

## What Each Method Actually Optimises

| Method | Objective |
|--------|-----------|
| Modularity (Louvain) | $\max_z \sum_c [m_c/m - (\kappa_c/2m)^2]$ |
| SBM MLE | $\max_{z,M} \ln \mathcal{L}(G \mid z, M)$ |
| Label propagation | Fixed point of $F = D^{-1}AF$ (with clamping) |
| GCN | $\min_W \mathcal{L}_{\text{CE}}(Z[V_L], y[V_L])$ |
| DeepWalk / Node2Vec | $\max_Z \sum_{(u,v) \in \text{walks}} \log \sigma(z_u^\top z_v)$ |
| Katz link prediction | $s(i,j) = [(I - \beta A)^{-1} - I]_{ij}$ |

---

## References

- Blondel, V. D. et al. (2008). Fast unfolding of communities in large networks.
  *Journal of Statistical Mechanics*, P10008.
- Traag, V. A., Waltman, L. & van Eck, N. J. (2019). From Louvain to Leiden.
  *Scientific Reports*, 9, 5233.
- Zhu, X. & Ghahramani, Z. (2002). Learning from labeled and unlabeled data
  with label propagation. *CMU Technical Report* CMU-CALD-02-107.
- Adamic, L. A. & Adar, E. (2003). Friends and neighbors on the web.
  *Social Networks*, 25(3), 211–230.
- Perozzi, B., Al-Rfou, R. & Skiena, S. (2014). DeepWalk. *KDD 2014*.
- Grover, A. & Leskovec, J. (2016). Node2Vec. *KDD 2016*.
- Kipf, T. N. & Welling, M. (2017). Semi-supervised classification with
  graph convolutional networks. *ICLR 2017*.
- Hamilton, W. L., Ying, R. & Leskovec, J. (2017). Inductive representation
  learning on large graphs. *NeurIPS 2017*.
- Veličković, P. et al. (2018). Graph attention networks. *ICLR 2018*.
- Xu, K. et al. (2019). How powerful are graph neural networks? *ICLR 2019*.
- Gilmer, J. et al. (2017). Neural message passing for quantum chemistry.
  *ICML 2017*.
- Newman, M. E. J. (2013). *Networks: An Introduction*. Ch. 11–14.
